---
title: "agriGrowthFlow 1.0.0: Cheatsheet EN"
subtitle: "Auditable workflow for plant growth analysis with persistent experimental unit"
author: "Walter E. Pereira and Magali H. Pereira Martinez"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: show
    code-tools: true
    embed-resources: true
    theme: cosmo
execute:
  freeze: auto
  warning: false
  message: false
bibliography: ../vignettes/references.bib
---

::: {.callout-note}
## Key concept

Declare the persistent experimental unit and sampling mechanism before fitting or interpreting any growth trajectory.
:::

# Domain question

How to compare treatments when measurements are repeated on the same plant or when harvests are destructive across different plots?

# Teaching data

All data are simulated and frozen for software testing, not field evidence.

```{r}
library(agriGrowthFlow)
maize <- growth_example_data("maize_destructive")
bean <- growth_example_data("bean_repeated")
sun <- growth_example_data("sunflower_sigmoid")
dim(maize); dim(bean); dim(sun)
```

# Workflow 1: qualitative factor with block

```{r}
gmaize <- growth_data(
  maize, time="day", sampling="destructive",
  experimental_unit="plot_id", treatment="nitrogen", block="block",
  total_mass="total_mass_g", leaf_area="leaf_area_m2"
)
growth_validate(gmaize)
growth_plan(gmaize)

cf <- growth_curve_coefficients(maize, time="day", response="total_mass_g",
  unit="plot_id", group="nitrogen", degree=2)
head(cf)

mv <- growth_manova(maize, time="day", response="total_mass_g",
  unit="plot_id", group="nitrogen", degree=2)
mv$multivariate
```

::: {.callout-tip}
## Study tip

Grouping letters are computed within each stratum; incomplete families are refused.
:::

# Workflow 2: quantitative factor with nonlinear regression

```{r}
one <- subset(sun, cultivar == unique(sun$cultivar)[1])
fit <- growth_fit(one, "logistic", time="day", response="biomass_g", n_start=5, seed=20260827)
fit
growth_traits(fit)
growth_predict(fit, time=seq(10,70,10), interval="confidence")
growth_diagnose(fit)
```

# Workflow 3:Quali + quanti (mixture)

```{r}
fit_set <- growth_fit(one, c("logistic","gompertz"), time="day", response="biomass_g",
  n_start=5, seed=20260827)
fit_set
growth_compare(fit_set)
```

# Workflow 4: model comparison with fit-quality table

```{r}
comp <- growth_compare(fit_set)
comp[, c("model","n","rmse","aic","aicc","delta_aicc","akaike_weight")]

# Grouped example: one fit per cultivar
fit_grp <- growth_fit(sun, "logistic", time="day", response="biomass_g",
  group="cultivar", n_start=3, seed=20260827)
fit_grp
```

::: {.callout-warning}
## Common mistake

Model comparison is valid only for identical prepared observations. The package refuses comparisons with different keys.
:::

# Longitudinal and flexible

```{r}
if (requireNamespace("nlme", quietly=TRUE)) {
  gbean <- growth_data(bean, time="day", sampling="repeated",
    experimental_unit="plant_id", treatment="water_regime", block="block",
    leaf_area="projected_leaf_area_m2")
  m <- growth_mixed(gbean, response="height_cm", degree=2, random_degree=1)
  print(m)
}

s <- growth_smooth(bean, time="day", response="height_cm", method="smooth_spline")
growth_smooth_predict(s, n=30)
growth_derivative(s, order=1, n=30)
```

# Bootstrap uncertainty

```{r}
one <- subset(sun, cultivar == unique(sun$cultivar)[1])
fit2 <- growth_fit(one, "logistic", time="day", response="biomass_g", n_start=3, seed=1)
b <- growth_boot(fit2, R=200, seed=20260827)
growth_boot_ci(b, source="coefficients")
```

::: {.callout-important}
## Critical information

Bootstrap must preserve the experimental hierarchy. `cluster` methods resample complete trajectories.
:::

# Consolidated workflow 1.0.0

```{r}
sun <- growth_example_data("sunflower_sigmoid")
plan <- growth_workflow(sun, time="day", response="biomass_g",
  strategy="parametric", models=c("logistic","gompertz"), n_start=5, seed=20260827)
growth_workflow_audit(plan)
growth_table(plan, "traits")
td <- tempfile("bundle_")
growth_export(plan, td, format="bundle")
list.files(td)
```

# Diagram

```{mermaid}
flowchart TD
  A[Declare persistent unit] --> B[Validate sampling]
  B --> C[Classical indices]
  C --> D{Parametric fit?}
  D -->|yes| E[Multi-start + diagnostics]
  D -->|no| F[Smooth / FPCA / MANOVA]
  E --> G[Biological traits]
  F --> G
  G --> H[Bootstrap / Bayes opt]
  H --> I[Audit + bundle]
```

# Best practices

- Do not treat destructive subsamples as longitudinal individuals.
- Diagnose covariance, heterogeneity and support before claims.
- Report conditional economic assumptions in `growth_harvest_opt()`.

# Exercises

::: {.callout-caution collapse="true"}
## Solutions

1. Redo workflow 2 with `n_start=10` and compare R².
2. Switch `growth_smooth` method to `loess` span 0.75 and evaluate derivative.
:::

# How to cite

```{r}
citation("agriGrowthFlow")
packageVersion("agriGrowthFlow")
R.version.string
```

```{r}
# dynamic numbers
fit <- growth_fit(subset(sun, cultivar==unique(sun$cultivar)[1]), "logistic", time="day", response="biomass_g", n_start=3, seed=1)
tr <- growth_traits(fit)
cat("Maximum rate:", round(tr$maximum_absolute_rate,2), "at", round(tr$time_maximum_rate,1), "days\n")
```
