---
title: "Classical Plant Growth Analysis: AGR, RGR, and NAR"
author: "agriGrowthFlow development team"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
bibliography: references.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{Classical Plant Growth Analysis: AGR, RGR, and NAR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(agriGrowthFlow)
```

# 1. Classical growth analysis as a set of biological rates

Classical plant growth analysis converts measurements at discrete harvests into rates and ratios. Hunt's treatment remains a useful foundation because it separates absolute change, proportional change, leaf-area efficiency and allocation [@Hunt1990].

# 2. Absolute growth rate

For dry mass $W$,

$$
AGR = \frac{W_2-W_1}{t_2-t_1}.
$$

```{r}
growth_agr(10, 18, 0, 4)
```

If mass is g plant$^{-1}$ and time is days, AGR is g plant$^{-1}$ day$^{-1}$. AGR can be negative when net dry mass decreases.

# 3. Relative growth rate

$$
RGR = \frac{\ln W_2-\ln W_1}{t_2-t_1}.
$$

```{r}
growth_rgr(10, 18, 0, 4)
```

RGR requires positive masses because of the logarithm. The result has dimensions of reciprocal time when the same mass unit is used at both harvests.

RGR answers a different question from AGR. A young plant can have a high RGR while adding little absolute mass, whereas a large mature plant can add more grams per day with a lower proportional rate.

# 4. Net assimilation rate

For total dry mass $W$ and leaf area $A$,

$$
NAR =
\frac{W_2-W_1}{t_2-t_1}
\frac{\ln A_2-\ln A_1}{A_2-A_1}.
$$

```{r}
growth_nar(
  w1 = 10, w2 = 18,
  a1 = 0.08, a2 = 0.14,
  t1 = 0, t2 = 4
)
```

When leaf areas are equal, the analytical expression has a removable singularity. The implementation uses the limit:

$$
NAR = AGR/A.
$$

```{r}
growth_nar(10, 18, 0.10, 0.10, 0, 4)
```

# 5. Why NAR is not simply photosynthesis

NAR is a whole-plant dry-mass accumulation rate per unit leaf area over an interval. It integrates photosynthesis, respiration, translocation, tissue turnover and the changing light environment. It should not be labelled as a direct gas-exchange measurement.

# 6. Use stable experimental units

```{r}
d <- growth_example_data("maize_destructive")
g <- growth_data(
  d, time = "day", sampling = "destructive",
  experimental_unit = "plot_id", treatment = "nitrogen", block = "block",
  total_mass = "total_mass_g", leaf_area = "leaf_area_m2",
  leaf_mass = "leaf_mass_g", ground_area = "ground_area_m2"
)

idx <- growth_indices(g)
head(idx$intervals[c("plot_id", "nitrogen", "block", "t1", "t2", "AGR", "RGR", "NAR")])
```

The formulas are applied after quadrats within each plot-date are summarized. This prevents arbitrary quadrat-to-quadrat pairing.

# 7. Ontogenetic drift

RGR commonly changes with plant age. A single RGR over an entire season can hide important biological transitions. The classical interval table keeps every interval explicit.

```{r}
aggregate(RGR ~ nitrogen + t1 + t2, idx$intervals, mean)
```

# 8. Classical versus functional approaches

The functional approach fits a smooth function to size or log size and differentiates it. Shipley and Hunt later discussed spline and LOESS smoothers that estimate trends and derivatives without imposing a single global parametric growth function [@ShipleyHunt1996]. Those smooth methods are planned for a later `agriGrowthFlow` release.

Poorter proposed a synthesis that smooths classical RGR values to reduce harvest-to-harvest fluctuation while retaining their interval basis [@Poorter1989]. Version 0.1.0 deliberately reports the unsmoothed foundational quantities so the raw information remains visible.

# 9. Edge cases and interpretation

## Zero or negative mass

RGR is undefined for non-positive mass. Do not add arbitrary constants merely to make the logarithm work without a scientific reason.

## Declining leaf area

The classical NAR expression remains mathematically usable for positive decreasing leaf area because both $\ln A_2-\ln A_1$ and $A_2-A_1$ are negative. Interpretation becomes increasingly dependent on the assumption that interval averages summarize a changing canopy adequately.

## Defoliation and abrupt biomass loss

Classical smooth-interval assumptions become problematic when known losses occur between harvests. Anten and Ackerly developed an iterative approach specifically for plants experiencing periodic biomass and leaf-area losses [@AntenAckerly2001]. That method is outside 0.1.0 and is planned for a later disturbance module.

# 10. Reporting

Report the two harvest dates, the mass and leaf-area basis, the experimental unit, whether the values were raw observations or within-unit summaries, and the exact units. Avoid reporting RGR or NAR without stating the interval over which it was calculated.

# References
