---
title: "Model Comparison, Diagnostics, and Prediction"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
bibliography: references.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{Model Comparison, Diagnostics, and Prediction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", warning = FALSE, message = FALSE)
library(agriGrowthFlow)
```

# Purpose

A nonlinear curve should be evaluated as a scientific model, not only as a smooth line through observations. Version 0.2.0 therefore separates three tasks:

1. model comparison on identical observations;
2. diagnostics of a selected fit;
3. prediction with clearly labeled local interval approximations.

# Build a candidate set

```{r}
d <- growth_example_data("sunflower_sigmoid")
d <- subset(d, cultivar == unique(d$cultivar)[1])
cs <- growth_fit(
  d,
  c("logistic", "gompertz"),
  time = "day",
  response = "biomass_g",
  n_start = 8,
  seed = 1701
)
cs
```

# Same-observation rule

Information criteria are only comparable when models describe the same response observations under the same likelihood basis.

```{r}
cmp <- growth_compare(cs)
cmp
```

The table includes:

- residual sum of squares;
- RMSE;
- MAE;
- AIC;
- corrected AIC;
- BIC;
- delta AICc;
- Akaike weight.

The package refuses comparison if prepared observation keys differ.

```{r eval=FALSE}
a <- growth_fit(subset(d, day <= 49), "logistic", time = "day", response = "biomass_g")
b <- growth_fit(d, "gompertz", time = "day", response = "biomass_g")
growth_compare(a, b)
```

That refusal prevents an apparently precise but invalid ranking.

# AICc is not the only scientific criterion

A smaller AICc indicates better estimated information trade-off within the candidate set. It does not prove that the biological mechanism is correct.

After ranking models, compare derived traits.

```{r}
growth_traits(cs)
```

If two models have similar AICc but imply substantially different t90 values, the late-growth conclusion is model-dependent and should be reported as such.

# Small-sample AICc

AICc requires enough observations relative to the likelihood parameter count. When that condition fails, the package returns `NA` for AICc rather than manufacturing a value.

Akaike weights are calculated only from finite AICc values.

# Diagnose the fitted curve

```{r}
f <- cs$fits$logistic
dg <- growth_diagnose(f)
dg
```

# Residual magnitude

```{r}
dg$summary[, c("rmse", "mae", "mean_residual", "descriptive_r_squared")]
```

The reported R-squared is descriptive. It should not be used as the primary criterion for choosing among nonlinear equations.

# Residual records

```{r}
head(dg$residuals)
```

These records preserve prepared time and experimental-unit identifiers so users can create residual-versus-time, residual-versus-fitted, and unit-specific diagnostic graphics.

# Unit-aware lag diagnostics

Serial plant-growth observations can be correlated within plots or plants. Ordinary NLS in 0.2.0 does not model that covariance.

The package nevertheless computes descriptive lag summaries without joining unrelated units.

```{r}
dg$summary[, c(
  "lag1_residual_correlation",
  "durbin_watson_descriptive",
  "lag_pairs",
  "lag_diagnostics_scope"
)]
```

These quantities are warnings and descriptive clues. They are not a correction for serial dependence.

# Parameter boundaries

```{r}
dg$boundaries
```

A coefficient close to its hard bound can indicate:

- insufficient time support;
- confounding among nonlinear parameters;
- an inappropriate model form;
- a start/bound specification that needs scientific reconsideration.

Do not silently widen a bound and continue until a desired number appears.

# Covariance conditioning

```{r}
dg$summary$covariance_condition_number
```

A large condition number suggests that the local covariance matrix contains strongly coupled parameter directions. Flexible shape parameters often increase this risk.

# Multistart audit

```{r}
dg$attempts
```

Convergence should be described by the distribution of fitting attempts, not only by the best attempt.

# Point prediction

```{r}
growth_predict(f)
```

By default, prediction uses the prepared observed time values.

# Regular prediction grid

```{r}
grid <- seq(min(f$data$.t), max(f$data$.t), length.out = 40)
p0 <- growth_predict(f, time = grid)
head(p0)
```

# Confidence interval for the mean curve

```{r}
pc <- growth_predict(
  f,
  time = grid,
  interval = "confidence",
  level = 0.95
)
head(pc)
```

The interval propagates the local coefficient covariance through the nonlinear mean function with a numerical gradient.

The numerical gradient respects hard parameter boundaries. If a central perturbation would leave the allowed parameter space, a one-sided finite difference is used.

# Prediction interval for a new response

```{r}
pp <- growth_predict(
  f,
  time = grid,
  interval = "prediction",
  level = 0.95
)
head(pp)
```

The prediction interval adds residual variance to local mean-curve variance. It is therefore wider than the corresponding confidence interval under the fitted ordinary-NLS assumptions.

# Interval limitations

The 0.2.0 intervals are not:

- bootstrap intervals;
- Bayesian credible intervals;
- nonlinear mixed-effects population intervals;
- plot-specific random-effect intervals;
- simultaneous confidence bands.

Those capabilities are deferred to later package versions rather than being mislabeled.

# Plot the model

```{r}
plot(f)
```

```{r}
plot(f, interval = "prediction")
```

```{r}
plot(f, interval = "none")
```

The plot is intended for inspection. Publication figures should state what interval is shown and should retain observed data wherever possible.

# Diagnostics and destructive sampling

The sunflower teaching dataset contains two destructive subsamples per plot and harvest. When declared through `growth_data()` with `sampling = "destructive"`, the automatic fitting preparation averages the two subsamples within persistent plot and time.

```{r}
sun <- growth_example_data("sunflower_sigmoid")
s1 <- subset(sun, cultivar == unique(sun$cultivar)[1])
g <- growth_data(
  s1,
  time = "day",
  sampling = "destructive",
  experimental_unit = "plot_id",
  total_mass = "biomass_g"
)
f2 <- growth_fit(g, "logistic", n_start = 5, seed = 3)
f2$aggregation
nrow(s1)
nrow(f2$data)
```

This does not turn harvested subsamples into repeated individuals. It preserves the persistent plot as the serial unit.

# Diagnostics and repeated non-destructive sampling

For repeated plant measurements, ordinary NLS also ignores within-plant dependence. Version 0.2.0 warns instead of silently treating those records as independent for inference.

```{r eval=FALSE}
bean <- growth_example_data("bean_repeated")
gb <- growth_data(
  bean,
  time = "day",
  sampling = "repeated",
  experimental_unit = "plant_id",
  plant = "plant_id",
  treatment = "water_regime",
  leaf_area = "projected_leaf_area_m2"
)
growth_fit(gb, "logistic", response = "projected_leaf_area_m2", group = "water_regime")
```

The correct treatment-inference layer for such data requires explicit random effects and covariance structure, which are deferred beyond 0.2.0.

# Comparison workflow

```{r}
cs2 <- growth_fit(
  d,
  c("logistic", "gompertz"),
  time = "day",
  response = "biomass_g",
  n_start = 10,
  seed = 201
)

growth_compare(cs2)

growth_diagnose(cs2$fits$logistic)
growth_diagnose(cs2$fits$gompertz)

growth_traits(cs2)
```

# Interpretation sequence

A defensible interpretation can follow this order:

1. confirm identical observations;
2. inspect convergence across starts;
3. inspect boundaries and covariance conditioning;
4. inspect residual pattern over time and within unit;
5. compare RSS, RMSE and MAE;
6. compare AICc/BIC within the candidate set;
7. compare derived biological traits;
8. identify extrapolated traits;
9. state that ordinary NLS does not represent serial experimental-unit covariance when relevant.

# What not to do

Do not compare models fitted after different missing-value deletion rules.

Do not select a curve from AICc alone when its asymptote is biologically implausible.

Do not call a local delta-method confidence interval a bootstrap interval.

Do not interpret residual lag correlation as a fitted AR(1) coefficient.

Do not use a descriptive R-squared as evidence that the experimental design has been represented correctly.

# Reporting template

A concise methods statement should identify:

- equation family;
- parameterization;
- bounded fitting engine;
- number of starts and seed;
- aggregation before fitting;
- model-comparison criteria;
- diagnostic checks;
- interval type;
- derived traits;
- longitudinal limitations.

# Next vignette

Continue with `v10-derived-growth-traits.Rmd` for model-independent biological summaries such as inflection, maximum rate, t10, t50, t90, and observed-support AUC.
