---
title: "Derived Biological Traits from Parametric Growth Curves"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
bibliography: references.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{Derived Biological Traits from Parametric Growth Curves}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", warning = FALSE, message = FALSE)
library(agriGrowthFlow)
```

# Why derived traits matter

The raw coefficients of different nonlinear equations are not directly comparable. A more useful biological question is often:

- when did maximum growth occur?
- how large was the maximum absolute rate?
- when was 50% or 90% of the final level reached?
- how long did the main growth transition last?
- what total fitted growth exposure occurred over the measured period?

`growth_traits()` converts model-specific coefficients into a common trait table.

# Example candidate set

```{r}
d <- growth_example_data("sunflower_sigmoid")
d <- subset(d, cultivar == unique(d$cultivar)[1])
cs <- growth_fit(
  d,
  c("logistic", "gompertz"),
  time = "day",
  response = "biomass_g",
  n_start = 8,
  seed = 501
)
```

```{r}
tr <- growth_traits(cs)
tr
```

# Standard columns

The common table contains:

- `asymptote`;
- `inflection_time`;
- `inflection_response`;
- `maximum_absolute_rate`;
- `time_maximum_rate`;
- `t10`;
- `t50`;
- `t90`;
- `auc_observed_support`;
- `support_min`;
- `support_max`.

A missing value is preferable to a fabricated biological quantity when a trait is undefined for a model.

# Inflection time

The inflection point marks a change in curvature. For standard monotone sigmoids it also corresponds to the maximum absolute growth rate.

```{r}
growth_inflection(cs)
```

For logistic growth,

\[
t_{\mathrm{inf}}=\mathrm{mid},\qquad W(t_{\mathrm{inf}})=A/2.
\]

For Gompertz growth,

\[
t_{\mathrm{inf}}=\mathrm{mid},\qquad W(t_{\mathrm{inf}})=A/e.
\]

The equal name `mid` therefore does not imply the same response fraction.

# Richards inflection

Under the package Richards parameterization [@Richards1959], `mid` remains the inflection time, but the response fraction depends on `shape`:

\[
\frac{W(t_{\mathrm{inf}})}{A}=(1+\nu)^{-1/\nu}.
\]

This provides a direct biological interpretation of the asymmetry parameter.

# Maximum absolute rate

```{r}
growth_maxrate(cs)
```

The units are response units per time unit.

If biomass is g per plant and time is days after emergence, maximum absolute rate has units g plant\(^{-1}\) day\(^{-1}\).

If biomass is g m\(^{-2}\), the corresponding rate is g m\(^{-2}\) day\(^{-1}\).

Always report the response basis.

# t10, t50, and t90

```{r}
tr[, c("model", "t10", "t50", "t90")]
```

These are times to reach 10%, 50%, and 90% of a finite upper level.

The transition duration

\[
D_{10-90}=t_{90}-t_{10}
\]

can be useful for comparing how quickly cultivars or treatments traverse the principal growth phase.

```{r}
tr$duration_10_90 <- tr$t90 - tr$t10
tr[, c("model", "duration_10_90")]
```

# Custom fractional targets

```{r}
f <- cs$fits$logistic
growth_time_to(f, fraction = c(0.10, 0.25, 0.50, 0.75, 0.90))
```

Fractional targets require a finite upper level.

# Absolute targets

Absolute targets can be more useful when the asymptote is uncertain or when a management threshold has a natural response scale.

```{r}
growth_time_to(f, target = c(50, 100, 150))
```

A target beyond the finite upper level returns an unavailable time rather than an impossible finite estimate.

# Zero targets

The mathematical time associated with zero depends on the curve family.

For logistic, Gompertz, Richards, and expolinear trajectories, zero is approached asymptotically toward negative infinity.

For Chapman-Richards, Weibull, and von Bertalanffy forms in the package, zero is assigned at the fitted `origin` and earlier.

For beta growth, zero occurs at the biological elapsed-time origin.

```{r}
growth_time_to(f, target = 0)
```

# AUC is limited to observed support

```{r}
tr[, c("auc_observed_support", "support_min", "support_max")]
```

The package does not integrate a sigmoid curve to infinity because such a quantity would be dominated by the arbitrary extension of a positive asymptote.

Instead,

\[
AUC_{obs}=\int_{t_{min}}^{t_{max}} \hat W(t)\,dt.
\]

This trait summarizes fitted response exposure over the actual measured time window.

Comparisons of AUC are meaningful only when the support interval and response units are comparable.

# Expolinear traits

```{r}
w <- growth_example_data("wheat_expolinear")
w <- subset(w, nitrogen == unique(w$nitrogen)[1])
f_exp <- growth_fit(w, "expolinear", time = "day", response = "biomass_g_m2")
growth_traits(f_exp)
```

Expolinear growth has no finite upper biomass asymptote. Consequently:

- `asymptote` is unavailable;
- t10, t50, and t90 relative to an asymptote are unavailable;
- maximum absolute rate equals `cm` as an asymptotic rate;
- `time_maximum_rate` is infinite because the derivative approaches `cm` rather than attaining it at a finite time.

Absolute targets remain available.

```{r}
growth_time_to(f_exp, target = c(200, 400, 600))
```

# Beta-growth traits

For the determinate beta function [@YinEtAl2003; @YinEtAl2003Erratum]:

- `wmax` is the final size;
- `tm` is the time of maximum absolute growth rate;
- `te` is the finite end of growth.

```{r eval=FALSE}
f_beta <- growth_multistart(
  d,
  "beta_growth",
  time = "day",
  response = "biomass_g",
  n_start = 40,
  seed = 61
)
growth_traits(f_beta)
```

A fitted `te` is scientifically convincing only when the experiment provides data close enough to the end of growth. A mathematically finite estimate is not equivalent to empirical support.

# Grouped descriptive traits

```{r}
sun <- growth_example_data("sunflower_sigmoid")
grp <- growth_fit(
  sun,
  "logistic",
  time = "day",
  response = "biomass_g",
  group = "cultivar",
  n_start = 5,
  seed = 19
)
gtr <- growth_traits(grp)
gtr
```

The output is always a data frame and includes `group` for grouped collections.

```{r}
gtr$duration_10_90 <- gtr$t90 - gtr$t10
gtr[, c("group", "inflection_time", "maximum_absolute_rate", "duration_10_90")]
```

These quantities describe separate curves. They do not by themselves constitute a joint statistical test among cultivars.

# Trait extrapolation

A trait can be mathematically defined but weakly supported by data.

Suppose the observed maximum time is 56 days and fitted t90 is 84 days. The t90 estimate is an extrapolation beyond measured support.

Check this explicitly:

```{r}
tr$inflection_inside_support <- with(tr, inflection_time >= support_min & inflection_time <= support_max)
tr$t90_inside_support <- with(tr, t90 >= support_min & t90 <= support_max)
tr[, c("model", "inflection_inside_support", "t90_inside_support")]
```

An agronomic report should distinguish interpolation from extrapolation.

# Trait uncertainty in 0.2.0

Version 0.2.0 provides point estimates for derived traits. It does not attach bootstrap or Bayesian uncertainty intervals to t90, maximum rate, or AUC.

This is a deliberate boundary of the release.

Nonlinear transformation of uncertain coefficients can produce asymmetric trait uncertainty. A later bootstrap layer should re-estimate the entire model within each resample rather than only transforming coefficient standard errors.

# Comparing biological conclusions across models

```{r}
growth_compare(cs)
growth_traits(cs)
```

A useful sensitivity analysis asks whether the biological conclusion changes when a scientifically plausible alternative curve is used.

Examples:

- Is cultivar ranking by t50 stable?
- Is maximum absolute rate similar under logistic and Gompertz fits?
- Does t90 change enough to alter a harvest recommendation?
- Does AUC ranking depend on model family?

# Reporting derived traits

A trait table should include units and model context.

Recommended columns include:

- model;
- estimate;
- response unit;
- time unit;
- observed support;
- whether the trait lies inside support;
- uncertainty method, once available.

# What not to do

Do not compare t50 from curves fitted to different biological time origins without transformation.

Do not report expolinear t90 relative to a nonexistent finite asymptote.

Do not call an extrapolated t90 an observed maturity date.

Do not interpret model-specific `mid` coefficients as identical biological fractions across all families.

Do not imply bootstrap or Bayesian uncertainty where version 0.2.0 provides only point traits.

# Integrated workflow

```{r}
d <- growth_example_data("sunflower_sigmoid")
d <- subset(d, cultivar == unique(d$cultivar)[1])

fit <- growth_fit(
  d,
  c("logistic", "gompertz"),
  time = "day",
  response = "biomass_g",
  n_start = 8,
  seed = 73
)

growth_compare(fit)
growth_traits(fit)
growth_inflection(fit)
growth_maxrate(fit)
```

# Next vignette

The complete analysis sequence is developed in `v11-parametric-growth-workflow.Rmd`, which connects experimental design, classical growth analysis, parametric fitting, diagnostics, comparison, prediction, and biological interpretation.
