---
title: "agriGrowthFlow 0.2.0 Foundations-to-Parametric Tutorial: From Experimental Units to Biological Curve Traits"
author: "agriGrowthFlow development team"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
bibliography: references.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{agriGrowthFlow 0.2.0 Foundations-to-Parametric Tutorial: From Experimental Units to Biological Curve Traits}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE
)
library(agriGrowthFlow)
```

# 1. Tutorial goals

This tutorial connects the complete 0.1.0 foundation to the new 0.2.0 parametric growth layer. It is written as a reproducible analysis narrative rather than a catalogue of isolated functions.

By the end, the reader should be able to:

1. identify the persistent experimental unit in destructive and repeated studies;
2. declare plant-growth roles with `growth_data()`;
3. validate the design before curve fitting;
4. decide whether a parametric curve is scientifically useful;
5. inspect the built-in equation registry;
6. generate automatic starting values;
7. fit bounded nonlinear models from multiple starts;
8. distinguish grouped descriptive curves from formal longitudinal inference;
9. compare candidate models fitted to identical observations;
10. diagnose convergence, bounds and residual pattern;
11. predict mean trajectories and new observations;
12. extract inflection, maximum rate, target times and AUC;
13. understand when the beta-growth and expolinear equations answer different crop-growth questions;
14. report limitations that version 0.2.0 deliberately does not conceal.

The package uses classical plant-growth foundations [@Hunt1990], design-aware reasoning for growth curves [@KeulsGarretsen1982], nonlinear agricultural modeling principles [@ArchontoulisMiguez2015], the Richards growth family [@Richards1959], expolinear crop growth [@GoudriaanMonteith1990], and the corrected beta growth function [@YinEtAl2003; @YinEtAl2003Erratum].

# Part I. The experiment comes before the curve

## 2. Three levels that must not be confused

A growth experiment may contain:

- biological observation units, such as individual plants or harvested quadrats;
- experimental units, such as plots or pots to which treatment is independently assigned;
- repeated time observations or destructive harvests nested inside those units.

The same CSV file can contain all three.

A nonlinear fitting function cannot reconstruct this hierarchy reliably from column names alone. That is why `agriGrowthFlow` begins with explicit role declaration.

## 3. Serial destructive sunflower example

Load the frozen simulated dataset.

```{r}
sun <- growth_example_data("sunflower_sigmoid")
dim(sun)
head(sun)
```

The dataset contains:

- three cultivars;
- four blocks;
- nine harvest times;
- two destructive subsamples per plot and time.

Check the structure.

```{r}
with(sun, table(cultivar, block))
with(sun, table(day))
length(unique(sun$plot_id))
```

The subsample identifier changes. `plot_id` persists through the growth series.

## 4. Declare the data roles

```{r}
gsun <- growth_data(
  sun,
  time = "day",
  sampling = "destructive",
  experimental_unit = "plot_id",
  plot = "plot_id",
  block = "block",
  treatment = "cultivar",
  total_mass = "biomass_g",
  units = list(
    time = "days after emergence",
    total_mass = "g per sampled plant-equivalent"
  ),
  metadata = list(
    dataset = "simulated teaching data",
    purpose = "parametric growth-curve examples"
  )
)
gsun
```

No observations are changed by `growth_data()`. The function creates a semantic layer around the original data.

## 5. Validate before fitting

```{r}
val <- growth_validate(gsun)
val
```

For destructive harvests, different sampled plants at successive times are not repeated measurements of one individual plant. A persistent plot or comparable higher-level unit is required to define a serial experimental trajectory. This principle follows the design logic discussed by Keuls and Garretsen [@KeulsGarretsen1982].

## 6. Ask the planning function

```{r}
plan <- growth_plan(gsun)
plan
```

The plan is intentionally conservative. It does not select a nonlinear equation automatically. It first checks which classical and model-based analyses are structurally possible.

## 7. Visualize observations before equations

```{r}
growth_plot(
  gsun,
  response = "biomass_g",
  group = "cultivar",
  summary = "both"
)
```

Inspect:

- monotonicity;
- evidence of a plateau;
- asymmetry;
- between-plot variability;
- unusual harvests;
- cultivar-specific shapes;
- whether the observed interval includes early and late phases.

A curve model should follow this inspection, not replace it.

# Part II. Retain the classical growth foundation

## 8. Why 0.2.0 does not replace AGR and RGR

Parametric curves and classical interval indices answer overlapping but different questions.

AGR over an interval is

$$
AGR=\frac{W_2-W_1}{t_2-t_1}.
$$

RGR is

$$
RGR=\frac{\log W_2-\log W_1}{t_2-t_1}.
$$

These are direct interval summaries. A model-derived instantaneous derivative is conditional on the fitted equation.

Both can be scientifically useful.

## 9. Classical indices on a compatible dataset

The sunflower example does not contain leaf area. Use the maize foundation dataset for a complete classical workflow.

```{r}
maize <- growth_example_data("maize_destructive")

gm <- growth_data(
  maize,
  time = "day",
  sampling = "destructive",
  experimental_unit = "plot_id",
  plot = "plot_id",
  block = "block",
  treatment = "nitrogen",
  total_mass = "total_mass_g",
  leaf_area = "leaf_area_m2",
  leaf_mass = "leaf_mass_g",
  root_mass = "root_mass_g",
  stem_mass = "stem_mass_g",
  reproductive_mass = "reproductive_mass_g",
  ground_area = "ground_area_m2"
)
```

Calculate interval indices.

```{r}
idx <- growth_indices(gm)
idx
```

The destructive subsamples are aggregated inside the persistent plot-time combination before intervals are created.

## 10. A nonlinear curve does not repair a wrong unit

Suppose each harvested plant were incorrectly treated as a repeated subject. A nonlinear function might still produce a smooth fitted line. The numerical smoothness would not make the data structure correct.

This is why 0.2.0 inherits every design gate from 0.1.0.

# Part III. The 0.2.0 model registry

## 11. Inspect available equations

```{r}
models <- growth_models()
models
```

The table is the authoritative registry for version 0.2.0.

## 12. Model families in practical language

The logistic curve is symmetric around half its asymptote.

The Gompertz curve is asymmetric and reaches its inflection below half of the asymptote.

The Richards curve adds a shape parameter [@Richards1959].

Chapman-Richards and Weibull provide additional flexible monotone forms.

The von Bertalanffy implementation imposes an exponent of three.

Beta growth represents determinate growth with a finite end time [@YinEtAl2003].

Expolinear growth represents transition from exponential to approximately linear crop accumulation [@GoudriaanMonteith1990].

## 13. Do not treat the registry as eight equally plausible choices

For each study, remove equations that conflict with known biology.

Examples:

- If the process is determinate and an explicit maturity time is central, beta growth deserves special consideration.
- If crop stand biomass is expected to continue in an approximately linear accumulation regime over the observed period, expolinear may be more meaningful than forcing a finite asymptote.
- If the observed trajectory is symmetric around half of final size, logistic can be parsimonious.
- If no late growth phase was observed, asymptote-heavy comparisons should be interpreted cautiously.

# Part IV. Focus on one cultivar to teach nonlinear fitting

## 14. Create a one-cultivar dataset

```{r}
lev <- unique(sun$cultivar)
sun_a <- subset(sun, cultivar == lev[1])
head(sun_a)
```

Declare it without a treatment role, because the current data contain only one cultivar.

```{r}
g_a <- growth_data(
  sun_a,
  time = "day",
  sampling = "destructive",
  experimental_unit = "plot_id",
  plot = "plot_id",
  block = "block",
  total_mass = "biomass_g",
  units = list(time = "days after emergence", total_mass = "g")
)
```

## 15. Inspect automatic starting values

```{r}
st_log <- growth_start(g_a, "logistic")
st_log
```

Starting values are proposals for the optimizer. They are not biological estimates.

## 16. Fit logistic growth

```{r}
fit_log <- growth_fit(
  g_a,
  model = "logistic",
  n_start = 8,
  seed = 20260827
)
fit_log
```

Because the dataset retains plot-level repeated harvest summaries, version 0.2.0 warns that ordinary nonlinear least squares does not model the within-plot correlation structure. The fitted curve is useful for description and model screening. Formal treatment inference awaits the mixed-effects layer.

## 17. Inspect what was aggregated

```{r}
fit_log$aggregation
nrow(sun_a)
nrow(fit_log$data)
head(fit_log$data)
```

Two destructive subsamples at the same plot and harvest time become one plot-time mean under `aggregate = "auto"`.

The four plots remain distinct.

## 18. Inspect fitted coefficients

```{r}
coef(fit_log)
```

For logistic growth:

- `asym` is the fitted upper level;
- `mid` is inflection time and t50;
- `scale` controls the time scale of transition.

These interpretations are specific to the package parameterization.

## 19. Inspect fitting attempts

```{r}
fit_log$attempts
```

Report the number of attempted and converged starts when numerical sensitivity matters.

# Part V. Compare a parsimonious candidate set

## 20. Logistic versus Gompertz

```{r}
candidates <- growth_fit(
  g_a,
  model = c("logistic", "gompertz"),
  n_start = 8,
  seed = 1501
)
```

Compare.

```{r}
comparison <- growth_compare(candidates)
comparison
```

## 21. Interpret the comparison table

Do not read only the first row.

Ask:

1. How large is delta AICc?
2. Are RMSE and MAE meaningfully different?
3. Do the models imply different inflection timing?
4. Does one model estimate an implausible asymptote?
5. Is the model choice stable to multistart fitting?
6. Are important traits interpolated or extrapolated?

## 22. Examine biological traits for both candidates

```{r}
traits_candidates <- growth_traits(candidates)
traits_candidates
```

A small AICc difference combined with a large difference in t90 means the late-growth conclusion is model-dependent.

That sensitivity belongs in the scientific interpretation.

# Part VI. Diagnostics before conclusions

## 23. Diagnose the logistic fit

```{r}
diag_log <- growth_diagnose(fit_log)
diag_log
```

The diagnostics include:

- RMSE;
- MAE;
- mean residual bias;
- descriptive R-squared;
- lag-one residual association;
- descriptive Durbin-Watson ratio;
- covariance condition number;
- parameters near hard bounds;
- all multistart attempts.

## 24. Unit-aware lag summaries

If a stable unit is present, residual pairs are formed within unit.

This prevents the residual at the last time of one plot from being treated as adjacent to the first residual of another plot.

```{r}
diag_log$summary[, c(
  "lag1_residual_correlation",
  "durbin_watson_descriptive",
  "lag_diagnostics_scope"
)]
```

These measures remain descriptive because the fitted NLS model does not include a serial covariance structure.

## 25. Inspect residual records

```{r}
head(diag_log$residuals)
```

A residual plot can be created from the returned data with standard R or ggplot2 if needed.

## 26. Inspect boundary and identifiability warnings

```{r}
diag_log$issues
```

A parameter near a boundary can signal weak support.

A large covariance condition number can indicate parameter trade-offs.

Neither problem should be solved automatically by hiding the diagnostic.

# Part VII. Predictions and intervals

## 27. Predict on the observed time scale

```{r}
pred_obs <- growth_predict(fit_log)
head(pred_obs)
```

## 28. Predict on a regular grid

```{r}
grid <- seq(
  min(fit_log$data$.t),
  max(fit_log$data$.t),
  length.out = 20
)
pred_grid <- growth_predict(fit_log, time = grid)
head(pred_grid)
```

## 29. Approximate confidence band

```{r}
pred_conf <- growth_predict(
  fit_log,
  time = grid,
  interval = "confidence",
  level = 0.95
)
head(pred_conf)
```

This band propagates the estimated coefficient covariance through the nonlinear mean function by a numerical delta-method gradient.

## 30. Approximate prediction band

```{r}
pred_new <- growth_predict(
  fit_log,
  time = grid,
  interval = "prediction",
  level = 0.95
)
head(pred_new)
```

A prediction band additionally includes residual variance for a new response under the ordinary-NLS model.

## 31. Do not confuse interval types

A parameter confidence interval, mean-curve confidence band, prediction interval, bootstrap interval and Bayesian credible interval answer different questions. The expanded tutorial pattern used in this package project treats this distinction explicitly.

Version 0.2.0 implements local delta-method curve intervals, not bootstrap uncertainty for derived traits.

# Part VIII. Plot the fitted model

## 32. Default plot

```{r}
plot(fit_log)
```

The graph combines raw prepared observations, fitted trajectory and approximate confidence band.

## 33. Prediction band

```{r}
plot(fit_log, interval = "prediction")
```

Use prediction bands only when a new-observation interpretation is appropriate.

## 34. No interval

```{r}
plot(fit_log, interval = "none")
```

A simple curve can be preferable for a figure focused on biological shape rather than uncertainty.

# Part IX. Translate coefficients into biological traits

## 35. Standard trait table

```{r}
tr_log <- growth_traits(fit_log)
tr_log
```

## 36. Inflection

```{r}
growth_inflection(fit_log)
```

For logistic growth, the inflection response equals half of the asymptote.

## 37. Maximum absolute growth rate

```{r}
growth_maxrate(fit_log)
```

If biomass is in grams and time in days, the rate is grams per day on the response basis used to construct the curve.

## 38. Time to fractional targets

```{r}
growth_time_to(
  fit_log,
  fraction = c(0.10, 0.25, 0.50, 0.75, 0.90)
)
```

## 39. Time to absolute targets

```{r}
growth_time_to(
  fit_log,
  target = c(50, 100, 150)
)
```

Absolute targets can be preferable when the upper asymptote is uncertain.

## 40. AUC over observed support

The AUC returned by `growth_traits()` integrates only from observed minimum to observed maximum time.

```{r}
tr_log[, c("auc_observed_support", "support_min", "support_max")]
```

It is not an integral to infinity.

# Part X. Richards flexibility without automatic preference

## 41. Why Richards is attractive

Richards introduced a flexible growth family that can vary the location of the inflection relative to the asymptote [@Richards1959].

This can represent asymmetry beyond logistic growth.

## 42. Why flexibility can be costly

The shape parameter can trade off with asymptote and time scale. If the experiment has few late observations, different parameter sets can produce nearly indistinguishable curves over the measured range.

Use more starts and inspect conditioning.

```{r eval=FALSE}
fit_rich <- growth_multistart(
  g_a,
  "richards",
  n_start = 50,
  seed = 8128
)
growth_diagnose(fit_rich)
growth_traits(fit_rich)
```

A simpler model remains preferable when the additional shape parameter is not supported or does not change the scientific conclusion materially.

# Part XI. Determinate growth with the beta function

## 43. Biological motivation

Many annual crops and individual organs have a finite growth period. Yin et al. designed the beta growth function so that final size and end time are explicit [@YinEtAl2003].

The parameters are:

- `wmax`: final size;
- `tm`: time of maximum growth rate;
- `te`: end of growth.

## 44. Corrected equation

The correct exponent uses $t_e$ in the numerator [@YinEtAl2003Erratum].

This is not a cosmetic correction. It changes the function.

## 45. Fit syntax

```{r eval=FALSE}
fit_beta <- growth_multistart(
  g_a,
  "beta_growth",
  n_start = 50,
  seed = 9211
)
coef(fit_beta)
growth_traits(fit_beta)
```

## 46. Time origin requirement

The beta model assumes growth begins at time zero under the implemented form.

If measurements are recorded as Julian dates, calendar dates, or days after an unrelated event, construct a biologically meaningful elapsed-time variable first.

## 47. Do not infer `te` from early growth alone

If the final observation occurs long before maturity, the model can mathematically estimate `te`, but the estimate depends heavily on model shape.

Treat such a result as extrapolation.

# Part XII. Expolinear crop growth

## 48. Load the wheat teaching data

```{r}
wheat <- growth_example_data("wheat_expolinear")
head(wheat)
```

The response is crop biomass per ground area.

## 49. Inspect treatment trajectories

```{r}
with(wheat, table(nitrogen, day))
```

## 50. Focus on one nitrogen level

```{r}
n0 <- unique(wheat$nitrogen)[1]
w0 <- subset(wheat, nitrogen == n0)
```

Declare the serial destructive structure.

```{r}
gw <- growth_data(
  w0,
  time = "day",
  sampling = "destructive",
  experimental_unit = "plot_id",
  plot = "plot_id",
  block = "block",
  total_mass = "biomass_g_m2",
  ground_area = "ground_area_m2",
  units = list(
    time = "days after emergence",
    total_mass = "g m^-2",
    ground_area = "m2"
  )
)
```

## 51. Fit expolinear growth

```{r}
fit_exp <- growth_fit(
  gw,
  "expolinear",
  n_start = 8,
  seed = 1203
)
fit_exp
```

## 52. Interpret expolinear coefficients

```{r}
coef(fit_exp)
```

Under the Goudriaan-Monteith formulation [@GoudriaanMonteith1990]:

- `rm` controls the early exponential relative-rate regime;
- `cm` is the late maximum absolute crop growth rate approached by the model;
- `tb` positions the transition in time.

## 53. Expolinear maximum rate

```{r}
growth_maxrate(fit_exp)
```

The rate approaches `cm` asymptotically. The model has no finite upper biomass asymptote.

## 54. Absolute target times remain meaningful

```{r}
growth_time_to(
  fit_exp,
  target = c(200, 400, 600)
)
```

Fractional asymptote targets do not apply.

# Part XIII. Describe cultivar curves without claiming a joint inferential test

## 55. Fit one logistic curve per cultivar

```{r}
fit_groups <- growth_fit(
  gsun,
  "logistic",
  group = "cultivar",
  n_start = 6,
  seed = 1907
)
fit_groups
```

## 56. Extract grouped traits

```{r}
traits_groups <- growth_traits(fit_groups)
traits_groups
```

## 57. Calculate a transparent duration

```{r}
traits_groups$duration_10_90 <- traits_groups$t90 - traits_groups$t10
traits_groups[, c(
  "group",
  "inflection_time",
  "maximum_absolute_rate",
  "duration_10_90"
)]
```

These are descriptive group-specific model summaries.

## 58. Why separate curves are not a treatment hypothesis test

Fitting each cultivar separately does not fit a common residual covariance structure or a shared parameter model. The standard errors from separate ordinary NLS fits do not automatically define a valid test of coefficient differences.

The 0.3.0 longitudinal layer will address shared hierarchical models.

# Part XIV. Compare model forms within a cultivar, not across different data

## 59. Correct comparison

```{r}
cs_a <- growth_fit(
  g_a,
  c("logistic", "gompertz"),
  n_start = 6
)
growth_compare(cs_a)
```

Both models use the same prepared observations.

## 60. Incorrect comparison

```{r eval=FALSE}
f_a <- growth_fit(
  subset(sun, cultivar == lev[1]),
  "logistic",
  time = "day",
  response = "biomass_g"
)

f_b <- growth_fit(
  subset(sun, cultivar == lev[2]),
  "gompertz",
  time = "day",
  response = "biomass_g"
)

growth_compare(f_a, f_b)
```

The function refuses this because the observations differ.

# Part XV. A decision workflow for real experiments

## 61. Gate 1: verify biological time

Ask whether zero has a biological meaning.

Examples:

- days after emergence;
- days after transplanting;
- thermal time since emergence;
- days after anthesis for grain filling.

Time re-expression changes the interpretation of parameters such as `mid`, `origin`, `tb`, `tm`, and `te`.

## 62. Gate 2: verify experimental unit

Write down the unit independently randomized to treatment.

Do not infer it from the number of rows.

## 63. Gate 3: determine sampling mechanism

Choose among:

- repeated non-destructive;
- serial destructive;
- mixed.

A destructive harvest cannot repeatedly observe the same harvested plant.

## 64. Gate 4: inspect trajectory support

Count distinct times and examine whether early, rapid and late phases are actually observed.

Four time points may be a minimum planning reference for simple within-phase polynomial characterization in the Keuls and Garretsen discussion [@KeulsGarretsen1982], but complex nonlinear parameters often need richer time support.

## 65. Gate 5: choose a biologically defensible candidate set

Do not begin with every model in the registry.

Document why each candidate is included.

## 66. Gate 6: inspect automatic starts

```{r eval=FALSE}
growth_start(g_a, "logistic")
growth_start(g_a, "gompertz")
```

## 67. Gate 7: fit from multiple starts

```{r eval=FALSE}
growth_multistart(g_a, "logistic", n_start = 30, seed = 42)
```

Record failures.

## 68. Gate 8: inspect parameter plausibility

Ask whether:

- asymptotes exceed plausible plant size;
- scale parameters imply unrealistically abrupt transitions;
- origin parameters are far outside meaningful time;
- beta `te` is plausible for maturity;
- expolinear `cm` has a realistic crop-growth-rate magnitude.

## 69. Gate 9: inspect residual and boundary diagnostics

```{r eval=FALSE}
growth_diagnose(fit_log)
```

## 70. Gate 10: compare models only on identical observations

```{r eval=FALSE}
growth_compare(candidates)
```

## 71. Gate 11: compare biological conclusions

```{r eval=FALSE}
growth_traits(candidates)
```

A model choice that materially changes the biological conclusion deserves sensitivity reporting.

## 72. Gate 12: identify the inferential level

If the scientific claim concerns treatment differences in longitudinal parameters, ordinary NLS curve screening is not the final model when repeated experimental units create dependence.

Stop before making a formal claim and move to the later hierarchical layer.

# Part XVI. Common mistakes and how the package responds

## 73. Treating destructive plants as repeated subjects

Package response: design validation flags the absence of a persistent unit or contradictory identifiers.

## 74. Passing incompatible vector lengths to classical formulas

Package response: 0.2.0 explicitly rejects ambiguous vector recycling rather than allowing R to recycle values silently.

```{r eval=FALSE}
growth_agr(c(1, 2, 3), c(2, 3), 0, 1)
```

## 75. Fitting negative biomass with built-in growth equations

Package response: the model layer rejects negative responses because the current built-in equations are parameterized for non-negative growth quantities.

## 76. Using beta growth with negative elapsed time

Package response: automatic start generation stops and asks for a biologically meaningful non-negative time scale.

## 77. Supplying unnamed starting values

Package response: fitting stops because parameter order should not be guessed.

## 78. Comparing models fitted to different observations

Package response: `growth_compare()` stops.

## 79. Asking expolinear for t50 of a finite asymptote

Package response: fractional asymptote time is undefined. Use an absolute target.

## 80. Treating confidence and prediction intervals as the same object

Package response: `growth_predict()` requires the interval type to be declared explicitly.

# Part XVII. Numerical controls for local validation

## 81. Logistic midpoint

For a logistic curve with $A=100$, midpoint 30 and scale 5:

$$
W(30)=50.
$$

This is checked in the automated test suite.

## 82. Gompertz inflection response

For $A=100$:

$$
W(t_m)=100/e.
$$

## 83. Richards shape-one reduction

With shape equal to one, the package Richards equation reduces exactly to the package logistic equation.

This identity is included as a regression test.

## 84. Beta endpoints

For valid beta parameters:

$$
W(0)=0,
$$

and

$$
W(t_e)=W_{max}.
$$

The test is especially important because the original article required a published correction to equation 11 [@YinEtAl2003Erratum].

## 85. Expolinear late slope

At sufficiently large time, the finite difference over one time unit approaches `cm`.

The static validation battery reproduces this known-truth check independently of R.

# Part XVIII. Reference integrity

## 86. Two-source metadata verification

The package keeps a machine-readable verification table under:

```text
inst/metadata/reference_verification.csv
```

For version 0.2.0, the added core records were checked as follows:

- Richards (1959): Oxford Academic and CiNii Research;
- Goudriaan and Monteith (1990): Oxford Academic and Wageningen University & Research;
- Yin et al. (2003): Oxford Academic and PubMed/PMC;
- Yin et al. erratum (2003): Oxford Academic and PubMed Central;
- Archontoulis and Miguez (2015): Agronomy Journal publisher record and an independent indexed scientific record.

The supplied PDFs are used to verify scientific content, but bibliographic metadata verification is kept as a separate audit step.

# Part XIX. What version 0.2.0 deliberately does not implement

## 87. Nonlinear mixed effects

Not yet in the parametric layer.

Ordinary NLS cannot replace a model with random plot or plant effects when formal longitudinal inference is required.

## 88. Residual covariance models

Not yet fitted by `growth_fit()`.

Lag diagnostics are descriptive.

## 89. Bootstrap uncertainty

Not yet applied to nonlinear curve coefficients or derived traits in the public 0.2.0 API.

It is planned for the uncertainty layer.

## 90. Bayesian nonlinear growth

Not yet in 0.2.0.

Hierarchical priors and posterior propagation will be added later rather than simulated here.

## 91. Multiphasic growth

Not yet in 0.2.0.

Diphasic logistic growth illustrates why inflection points of a sum of logistic components are not necessarily the component inflection points and may require numerical derivative roots [@MischanEtAl2015]. That belongs to the dedicated multiphasic module rather than being approximated in the current registry.

## 92. Smooth derivatives and functional analysis

Spline, GAM and functional-data derivatives are planned for the flexible-growth layer.

# Part XX. Reporting templates

## 93. Methods template for a descriptive curve screen

A methods section should state, in substance:

> Growth trajectories were defined at the persistent experimental-unit level. Destructive subsamples were averaged within experimental unit and harvest time before curve fitting. Biologically plausible candidate nonlinear equations were fitted by bounded nonlinear least squares using deterministic initialization followed by multiple admissible starting values. Candidate models were compared only on identical observations using AICc together with residual error, convergence, boundary diagnostics and biological plausibility. Derived traits included inflection time, maximum absolute growth rate and prespecified target times. Because ordinary nonlinear least squares does not represent within-unit serial dependence, this stage was used for trajectory description and model screening rather than final treatment inference.

Adapt this text to the actual experiment and do not copy methods that were not run.

## 94. Results template

A results paragraph should emphasize patterns:

> Both candidate sigmoid equations described the observed accumulation phase, but their late-growth implications differed. The selected curve placed the period of maximum absolute accumulation within the observed rapid-growth interval and produced target-time estimates consistent with the measured trajectory. Late asymptotic traits were more model-sensitive than inflection timing, so conclusions about final size were interpreted cautiously.

Insert actual estimates, uncertainty and units from the real analysis.

## 95. Figure caption template

> Observed plant growth and fitted nonlinear trajectory. Points represent experimental-unit summaries at each sampling time; the line is the selected parametric mean curve. The shaded region is the local 95% confidence band from the ordinary nonlinear least-squares coefficient covariance matrix and does not incorporate random experimental-unit effects.

# Part XXI. A compact end-to-end script

## 96. Data declaration

```{r eval=FALSE}
d <- growth_example_data("sunflower_sigmoid")
a <- subset(d, cultivar == unique(d$cultivar)[1])

g <- growth_data(
  a,
  time = "day",
  sampling = "destructive",
  experimental_unit = "plot_id",
  block = "block",
  total_mass = "biomass_g"
)
```

## 97. Validation

```{r eval=FALSE}
growth_validate(g)
growth_plan(g)
```

## 98. Candidate fitting

```{r eval=FALSE}
cs <- growth_fit(
  g,
  c("logistic", "gompertz"),
  n_start = 20,
  seed = 20260827
)
```

## 99. Comparison

```{r eval=FALSE}
growth_compare(cs)
```

## 100. Diagnostics

```{r eval=FALSE}
growth_diagnose(cs$fits$logistic)
growth_diagnose(cs$fits$gompertz)
```

## 101. Biological interpretation

```{r eval=FALSE}
growth_traits(cs)
```

## 102. Prediction

```{r eval=FALSE}
best <- cs$fits[[1]]
growth_predict(
  best,
  time = seq(0, 80, by = 5),
  interval = "confidence"
)
```

# Part XXII. Final checklist

## 103. Before accepting a fitted curve

Confirm all of the following:

- [ ] experimental unit is explicitly identified;
- [ ] destructive versus repeated sampling is correctly declared;
- [ ] time has a biological interpretation;
- [ ] response units are recorded;
- [ ] raw trajectories were inspected;
- [ ] candidate equations were chosen for scientific reasons;
- [ ] automatic starts were inspected or at least recorded;
- [ ] multiple starts were used when needed;
- [ ] failed attempts were retained;
- [ ] parameter bounds were inspected;
- [ ] covariance conditioning was checked;
- [ ] model comparison used identical observations;
- [ ] derived traits were compared among plausible models;
- [ ] extrapolated traits were identified;
- [ ] confidence and prediction intervals were not confused;
- [ ] ordinary NLS was not presented as a mixed longitudinal analysis;
- [ ] the beta equation uses the published corrected exponent;
- [ ] expolinear fractional asymptote targets were not invented;
- [ ] references and metadata were verified.

# 104. Final perspective

Version 0.2.0 extends `agriGrowthFlow` from classical plant growth analysis into biologically interpretable nonlinear trajectories while preserving the design-first philosophy of version 0.1.0.

The most important addition is not the number of equations. It is the workflow around them:

```text
experimental design
      |
      v
role declaration
      |
      v
design validation
      |
      v
raw trajectory inspection
      |
      v
biological candidate models
      |
      v
automatic starts and bounds
      |
      v
multistart nonlinear fitting
      |
      v
comparison and diagnostics
      |
      v
predictions and biological traits
      |
      v
explicit limitations before inference
```

The next modeling layer can now add nonlinear mixed effects, structured residual covariance and flexible smooth trajectories without rebuilding the foundations or weakening the experimental-unit safeguards established here.
