---
title: "Bayesian Diagnostics, PSIS-LOO, and Growth-Model Averaging"
author: "agriGrowthFlow"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
bibliography: references.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{Bayesian Diagnostics, PSIS-LOO, and Growth-Model Averaging}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", warning = FALSE, message = FALSE)
library(agriGrowthFlow)
```

# 1. Three questions that must remain separate

A Bayesian growth analysis contains at least three distinct evaluation problems:

1. Did the Markov chains explore the intended posterior distribution adequately?
2. Does the fitted model generate data resembling scientifically relevant features of the observations?
3. How well is the model expected to predict held-out observations compared with plausible alternatives?

Version 0.4.0 uses `posterior`-based MCMC diagnostics, `brms` posterior predictive checks, and `loo` Pareto-smoothed leave-one-out cross-validation. These tools are complementary rather than interchangeable.

The modern rank-normalized \(\hat R\) and effective sample-size framework is described by @VehtariEtAl2021. PSIS-LOO methodology is developed by @VehtariGelmanGabry2017. Model stacking follows the predictive-combination framework of @YaoEtAl2018.

# 2. Fit competing models to identical observations

The critical requirement for predictive comparison is that candidate models use the same prepared observations. Missingness filtering, aggregation, or group subsets must not differ silently between models.

```{r setup-data}
sun <- growth_example_data("sunflower_sigmoid")
sun1 <- subset(sun, cultivar == unique(sun$cultivar)[1])
```

```{r fit-models, eval=FALSE}
b_log <- growth_bayes(
  sun1,
  "logistic",
  time = "day",
  response = "biomass_g",
  random = "none",
  group_effects = "none",
  chains = 4,
  iter = 2500,
  warmup = 1250,
  seed = 5001
)

b_gom <- growth_bayes(
  sun1,
  "gompertz",
  time = "day",
  response = "biomass_g",
  random = "none",
  group_effects = "none",
  chains = 4,
  iter = 2500,
  warmup = 1250,
  seed = 5002
)
```

`agri_growth_bayes` objects store an `observation_key`. `growth_model_average()` refuses to combine candidate models when these keys differ.

# 3. Gate 1: sampler diagnostics

```{r diag, eval=FALSE}
d_log <- growth_bayes_diagnose(b_log)
d_gom <- growth_bayes_diagnose(b_gom)

d_log
d_gom
```

A predictive score should not be used to choose between poorly sampled posterior distributions. First resolve divergences, \(\hat R\) problems, or inadequate ESS.

# 4. Rank-normalized R-hat

For multiple chains, \(\hat R\) compares between-chain and within-chain behavior. Values near one are desirable. The 0.4.0 diagnostic wrapper flags values above 1.01, following current conservative practice described by @VehtariEtAl2021.

A passing \(\hat R\) is necessary but not sufficient. It does not guarantee that the scientific model is adequate.

# 5. Bulk and tail ESS

Bulk ESS concerns central posterior summaries such as means and medians. Tail ESS is particularly relevant to interval endpoints and tail probabilities.

```{r diag-table, eval=FALSE}
subset(d_log$parameters, rhat > 1.01 | ess_bulk < 400 | ess_tail < 400)
```

When the scientific result depends on a narrow tail probability, a low tail ESS deserves more attention than the raw total number of stored draws.

# 6. Divergences

Hamiltonian Monte Carlo divergences indicate that numerical trajectories encountered problematic posterior geometry. A model can have \(\hat R\) near one and still have divergences.

`growth_bayes_diagnose()` reports their count explicitly.

Potential responses include:

- reconsidering parameterization;
- using more informative but scientifically justified priors;
- simplifying a weakly identified hierarchy;
- increasing `adapt_delta` when appropriate;
- collecting more informative data.

# 7. Gate 2: posterior predictive checking

```{r ppcheck, eval=FALSE}
growth_pp_check(b_log, type = "dens_overlay", ndraws = 100)
growth_pp_check(b_log, type = "scatter_avg", ndraws = 100)
```

A density overlay checks marginal response distribution but can miss temporal structure. A growth analysis should also inspect predictions across time and, when relevant, by treatment or persistent unit.

The central question is not whether simulated datasets look visually identical to observed data. It is whether discrepancies relevant to the scientific question remain unexplained.

# 8. Gate 3: predictive evaluation with PSIS-LOO

Leave-one-out cross-validation asks how well a model predicts each observation when that observation is left out of its fitting information. Exact refitting for every observation is expensive. PSIS-LOO approximates this process from one posterior fit and diagnoses when importance sampling becomes unreliable [@VehtariGelmanGabry2017].

```{r loo, eval=FALSE}
l_log <- growth_loo(b_log)
l_gom <- growth_loo(b_gom)

l_log
l_gom
```

The returned object retains the complete `psis_loo` result and Pareto-k values.

# 9. Pareto-k diagnostics

The importance ratios used by PSIS-LOO can become unstable for influential observations. `growth_loo()` reports counts above 0.7 and above 1.

```{r pareto, eval=FALSE}
summary(l_log$pareto_k)
which(l_log$pareto_k > 0.7)
```

Large Pareto-k values are not merely a nuisance to hide. They can reveal influential harvests, underrepresented phases of growth, or model-data conflict.

# 10. Moment matching

For some difficult observations, `loo` can use moment matching to improve the PSIS approximation.

```{r moment, eval=FALSE}
l_log_mm <- growth_loo(b_log, moment_match = TRUE)
```

Moment matching should be documented when used. If diagnostics remain problematic, exact leave-one-out refitting or K-fold cross-validation may be needed outside the convenience wrapper.

# 11. LOO is predictive, not ontological

A lower expected predictive loss does not prove that one curve is the true biological law. Candidate growth equations are approximations. A model can predict well over observed time while giving less credible extrapolation or less interpretable parameters.

Model comparison should therefore combine:

- predictive performance;
- residual and posterior predictive behavior;
- parameter identifiability;
- biological plausibility;
- uncertainty in derived traits.

# 12. Why identical observations are enforced

If one candidate uses 80 observations and another 74 because of different preprocessing, their pointwise log predictive densities are not directly comparable in the intended way.

Version 0.4.0 stores a prepared-data key in each Bayesian object and refuses model averaging when candidate keys differ.

This is a deliberate safety feature.

# 13. Model stacking

When several models make complementary predictions, selecting one winner can discard useful predictive information. Stacking chooses nonnegative weights that optimize the predictive performance of the weighted predictive distribution [@YaoEtAl2018].

```{r stacking, eval=FALSE}
avg_w <- growth_model_average(
  b_log,
  b_gom,
  method = "stacking",
  type = "weights"
)

avg_w
```

The weights sum to one. They are predictive combination weights, not posterior probabilities that a model is true.

# 14. Pseudo-BMA and pseudo-BMA+

The alternative `method = "pseudobma"` uses LOO-based model scores. With `BB = TRUE`, the `loo` implementation uses Bayesian-bootstrap uncertainty for pseudo-BMA+ weights.

```{r pseudobma, eval=FALSE}
growth_model_average(
  b_log,
  b_gom,
  method = "pseudobma",
  BB = TRUE,
  type = "weights"
)
```

Stacking and pseudo-BMA answer similar predictive-combination questions with different optimization principles. The choice should be stated in the Methods section.

# 15. Model-averaged expected response

Weights can be used to form a mixture of posterior expected response draws.

```{r epred, eval=FALSE}
avg_epred <- growth_model_average(
  b_log,
  b_gom,
  method = "stacking",
  type = "epred",
  ndraws = 2000,
  seed = 5010
)

head(avg_epred$prediction)
```

`type = "epred"` averages uncertainty in the conditional mean response. It does not add a new observation-level residual draw.

# 16. Model-averaged predictive distribution

For future observations, use `type = "predict"`:

```{r predict, eval=FALSE}
avg_pred <- growth_model_average(
  b_log,
  b_gom,
  method = "stacking",
  type = "predict",
  ndraws = 2000,
  seed = 5011
)
```

This includes observation-level predictive variability from the individual Bayesian models.

# 17. Expected response versus future observation

This distinction should be explicit in figures:

- expected-response interval: uncertainty about the fitted mean trajectory;
- predictive interval: uncertainty for a new response, including residual variation.

A predictive interval is normally wider.

# 18. How the mixture is constructed

Version 0.4.0 uses a Monte Carlo mixture:

1. draw a candidate model according to the LOO weight vector;
2. sample one posterior expected or predictive trajectory from that model;
3. repeat for the requested number of mixture draws;
4. summarize the resulting mixture distribution pointwise.

This respects the full predictive mixture instead of merely averaging model-specific interval endpoints.

# 19. New time grids

A model-averaged prediction can be requested on a common standardized `newdata` grid.

```{r newdata, eval=FALSE}
nd <- b_log$data[rep(1, 101), , drop = FALSE]
nd$agf_time <- seq(min(b_log$data$agf_time), max(b_log$data$agf_time), length.out = 101)

avg_grid <- growth_model_average(
  b_log,
  b_gom,
  type = "epred",
  newdata = nd,
  ndraws = 2000,
  seed = 5012
)
```

Do not silently extrapolate the grid beyond the observed support when the manuscript claims empirical inference within the experiment.

# 20. Model averaging does not solve model misspecification

If all candidate models omit an important stress response or phase change, their weighted average still omits it. Model averaging distributes weight among the submitted candidates. It does not create a biologically missing candidate.

# 21. Relationship to bootstrap selection stability

Bootstrap selection stability from `growth_selection_stability()` and Bayesian stacking answer different questions.

Bootstrap stability asks:

> How often does each model win a selection criterion when the observed sample is resampled?

Stacking asks:

> Which predictive weights combine posterior predictive distributions to perform well under LOO?

A model can have moderate bootstrap selection frequency and still receive substantial stacking weight.

# 22. Model-specific traits versus averaged predictions

Model averaging is straightforward for predictions on a common response scale. Averaging a model-specific parameter such as a Logistic scale with a Gompertz scale is not meaningful.

For biological traits with common definitions, such as time to 50% of each model's asymptote, one can compare model-specific posterior trait distributions. A future package version may add trait-level model mixtures after the common estimand is formally defined for all included families.

# 23. A complete evaluation sequence

For each Bayesian candidate:

1. inspect prior predictive behavior;
2. fit the posterior model;
3. verify \(\hat R\), ESS, and divergences;
4. inspect posterior predictive behavior;
5. compute PSIS-LOO;
6. inspect Pareto-k;
7. compare predictive performance;
8. evaluate biological trait sensitivity;
9. use stacking only if a predictive mixture serves the scientific goal.

# 24. What not to do

Do not:

- compute LOO from a model whose sampler diagnostics fail;
- ignore Pareto-k values because the total LOO score looks favorable;
- compare models fitted to different observations;
- call stacking weights model probabilities;
- average coefficients with incompatible meanings across curve families;
- use predictive model averaging to hide poor candidate specification;
- report only the best LOO score without uncertainty or diagnostics.

# 25. Methods-language template

> Bayesian candidate growth models were fitted to identical prepared observations and evaluated using Pareto-smoothed leave-one-out cross-validation. Pareto-k diagnostics were inspected before model comparison. Predictive model weights were obtained by stacking of pointwise leave-one-out predictive densities. Population-level expected-response summaries were generated from a Monte Carlo mixture of model-specific posterior predictions using the estimated stacking weights.

State whether `epred` or `predict` was used, and report software versions.

# 26. Results-language template

> Both candidate models satisfied the prespecified MCMC diagnostics. PSIS-LOO diagnostics were examined for influential observations before computing predictive weights. The resulting stacking distribution was used to summarize the expected growth trajectory while preserving model-form uncertainty among the submitted candidates.

Again, numerical values must come from the actual final fit.

# 27. Reproducibility checklist

Record:

- exact candidate models;
- identical-data check;
- priors;
- posterior diagnostics;
- LOO version;
- Pareto-k diagnostics;
- weighting method;
- Bayesian bootstrap option for pseudo-BMA;
- number of mixture draws;
- random-effects setting in predictions;
- prediction grid;
- seed.

# 28. Summary

Version 0.4.0 treats Bayesian model evaluation as a gated process:

**sampler diagnostics -> posterior predictive checking -> PSIS-LOO diagnostics -> predictive comparison -> optional model averaging.**

This sequence prevents a superficially convenient predictive score from bypassing computational or scientific model checks.

# References
