---
title: "agriGrowthFlow 1.0.0: Reporting, Audit, and Reproducible Export"
author: "agriGrowthFlow team"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{agriGrowthFlow 1.0.0: Reporting, Audit, and Reproducible Export}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
link-citations: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(agriGrowthFlow)
```

# Purpose

A statistically correct model is not enough if the analysis cannot be reconstructed.

Version 1.0.0 therefore treats reporting and export as part of the analysis workflow.

This vignette explains:

* `growth_workflow_audit()`;
* `growth_table()`;
* `growth_report()`;
* `growth_export()`;
* the difference between a complete archive and a review table;
* how to package model output for collaborators and reviewers;
* what should and should not be written automatically.

# 1. Why reporting belongs inside the workflow

Plant growth analysis can combine destructive sampling, repeated measurements, nonlinear curves, smoothing, bootstrap, and Bayesian models.

Each layer introduces decisions.

A final table without those decisions is incomplete.

A reproducible analysis record should preserve both numerical results and the contract that generated them.

This is particularly important when the experimental unit is not the individual harvested plant [@KeulsGarretsen1982].

# 2. Create a planning workflow

```{r plan}
d <- growth_example_data("sunflower_sigmoid")
d1 <- subset(d, cultivar == unique(cultivar)[1])

w_plan <- growth_workflow(
  d1,
  time = "day",
  response = "biomass_g",
  objective = "trajectory",
  execute = FALSE,
  seed = 2001
)
```

A planning workflow already contains useful information.

```{r plan-names}
names(w_plan)
```

It records the objective, roles, method guide, selected strategy, decision ledger, seed, and execution status.

# 3. Audit the planning workflow

```{r audit-plan}
audit_plan <- growth_workflow_audit(w_plan)
audit_plan
```

The audit uses three statuses.

## PASS

The expected release-level gate is present.

## INFO

The item is optional, intentionally absent, or not applicable.

## FAIL

A structural problem blocks interpretation of the workflow as a complete analysis record.

A planning-only workflow should contain INFO records for fitting-related gates.

That is expected behavior.

# 4. Execute a parametric workflow

```{r run-workflow}
w <- growth_workflow(
  d1,
  time = "day",
  response = "biomass_g",
  strategy = "parametric",
  models = c("logistic", "gompertz"),
  n_start = 5,
  seed = 2002
)
w
```

Now audit it again.

```{r audit-run}
audit_run <- growth_workflow_audit(w)
audit_run
```

The audit does not say whether the model is biologically correct.

It says whether key workflow components are available for review.

# 5. A release audit is not a model diagnostic

The distinction is important.

A workflow audit asks:

* was a core model stored?;
* were diagnostics stored?;
* was a candidate comparison stored?;
* was uncertainty attached?;
* was a seed recorded?;

A model diagnostic asks different questions:

* did optimization converge?;
* are parameters near hard bounds?;
* is residual dependence visible?;
* is the covariance matrix ill-conditioned?;
* are Bayesian chains mixing adequately?;

Do not substitute one for the other.

# 6. Standardized tables

`growth_table()` provides compact views of common objects.

The default workflow view is the audit.

```{r table-auto}
growth_table(w)
```

Request traits explicitly.

```{r table-traits}
traits <- growth_table(w, component = "traits")
traits
```

Request candidate comparison.

```{r table-comparison}
comparison <- growth_table(w, component = "comparison")
comparison
```

Request the method guide.

```{r table-guide}
head(growth_table(w, component = "guide"))
```

# 7. Direct fit objects can also be tabulated

```{r direct-fit}
f <- growth_fit(
  d1,
  "logistic",
  time = "day",
  response = "biomass_g",
  n_start = 3,
  seed = 2003
)
```

```{r direct-coefficients}
growth_table(f, component = "coefficients")
```

The coefficient table is intentionally simple.

Use the original fit object for full covariance matrices, optimization records, and specialized methods.

# 8. Classical interval tables

The same interface works with classical growth analysis.

```{r classical-setup}
maize <- growth_example_data("maize_destructive")
gm <- growth_data(
  maize,
  time = "day",
  sampling = "destructive",
  experimental_unit = "plot_id",
  treatment = "nitrogen",
  block = "block",
  total_mass = "total_mass_g",
  leaf_area = "leaf_area_m2",
  leaf_mass = "leaf_mass_g",
  ground_area = "ground_area_m2"
)
ci <- growth_classical(gm)
```

```{r classical-table}
head(growth_table(ci, component = "intervals"))
```

These intervals preserve the persistent plot before rates are calculated.

Classical indices such as RGR, NAR, LAR, and SLA should retain their biological units and interpretation [@Hunt1990].

# 9. Diagnostics can be standardized when they are not nested

```{r direct-diagnostics}
diag_f <- growth_diagnose(f)
growth_table(diag_f, component = "diagnostics")
```

Nested diagnostics are not silently flattened.

If a workflow contains several candidate models, extract one diagnostic object explicitly.

```{r nested-diagnostics}
growth_table(w$diagnostics$logistic, component = "diagnostics")
```

This design avoids presenting model-specific diagnostics as if they belonged to one pooled model.

# 10. Markdown reporting

Generate report text in memory.

```{r report-memory}
report_text <- growth_report(w, title = "Sunflower growth analysis")
cat(paste(head(report_text, 20), collapse = "\n"))
```

The report contains:

* the release version;
* the objective;
* the selected strategy;
* execution status;
* uncertainty status;
* seed;
* workflow audit;
* candidate comparison when available;
* biological traits when available;
* workflow notes;
* interpretation safeguards.

# 11. What the report deliberately does not write

`growth_report()` does not automatically state that:

* one treatment is superior;
* one model is biologically true;
* a causal effect has been established;
* a harvest date should be adopted commercially;
* residual normality has been proven;
* a Bayesian model converged merely because it completed sampling.

These statements require scientific interpretation.

Automatic prose should not create claims that were not supported by the analysis.

# 12. Write the Markdown report to disk

```{r report-file}
report_file <- tempfile(fileext = ".md")
growth_report(w, file = report_file, title = "Sunflower growth analysis")
file.exists(report_file)
```

The Markdown file is intentionally portable.

It can be inspected without R.

It can be version controlled.

It can be converted later using Quarto or R Markdown if desired.

# 13. RDS is the complete archive

```{r rds-export}
rds_file <- tempfile(fileext = ".rds")
growth_export(w, rds_file, format = "rds")
file.exists(rds_file)
```

RDS preserves the complete nested object.

That includes fitted models, diagnostics, candidate comparisons, traits, workflow metadata, and uncertainty objects when attached.

For a complete R analysis archive, this is the preferred release format.

# 14. CSV is a review format

```{r csv-export}
csv_file <- tempfile(fileext = ".csv")
growth_export(audit_run, csv_file, format = "csv")
file.exists(csv_file)
```

CSV is useful for:

* collaborators who do not use R;
* spreadsheet review;
* supplementary tables;
* lightweight inspection;
* version-control diffs.

CSV cannot preserve a fitted nonlinear model.

It cannot preserve random-effects structures.

It cannot preserve bootstrap fit objects.

It cannot preserve Stan posterior draws and diagnostics as one nested analysis object.

# 15. Bundle export

The bundle combines the complete archive with human-readable review files.

```{r bundle-export}
bundle_dir <- tempfile(pattern = "agf_bundle_")
growth_export(w, bundle_dir, format = "bundle")
list.files(bundle_dir)
```

A workflow bundle contains at least:

* `object.rds`;
* `workflow_audit.csv`;
* `method_guide.csv`;
* `report.md`.

When available, it also contains:

* `traits.csv`;
* `candidate_comparison.csv`.

# 16. Why the bundle uses both RDS and flat files

There are two different reproducibility needs.

## Computational reproducibility

The complete object is required.

RDS serves this role.

## Review reproducibility

Humans often need to inspect a small number of important outputs quickly.

CSV and Markdown serve this role.

The bundle keeps both layers together.

# 17. Exporting uncertainty workflows

For bootstrap or Bayesian workflows, the complete object becomes even more important.

Bootstrap inference depends on the original fit, resampling method, successful replicate count, and empirical distribution of traits [@DavisonHinkley1997].

Bayesian inference depends on priors, posterior draws, chain diagnostics, predictive checks, and often PSIS-LOO diagnostics [@Buerkner2017; @VehtariGelmanGabry2017; @VehtariEtAl2021].

Do not reduce that information to one CSV file and discard the fitted object.

# 18. Reproducibility metadata

At minimum, archive:

* package version;
* R version;
* operating system;
* dependency versions;
* random seed;
* source-data checksum when possible;
* analysis script;
* complete workflow RDS;
* report;
* validation log.

For local release validation, also archive the exact package tarball that passed `R CMD check --as-cran`.

# 19. Recommended project structure

A simple project can use:

```text
analysis/
  01_prepare_data.R
  02_growth_models.R
  03_uncertainty.R
  04_report.R
outputs/
  agriGrowthFlow_bundle/
  figures/
  tables/
data/
  raw/
  processed/
validation/
  sessionInfo.txt
  package_check/
```

The exact structure can vary.

The important requirement is that raw data, analysis code, and derived outputs are not confused with each other.

# 20. Source-data checksums

Checksums help detect accidental changes in frozen inputs.

The package itself uses SHA-256 checksums for simulated teaching datasets and frozen source snapshots.

A research project can do the same for raw input files.

For example, from a shell:

```text
sha256sum data/raw/my_growth_data.csv
```

On Windows PowerShell:

```text
Get-FileHash data\raw\my_growth_data.csv -Algorithm SHA256
```

Record the hash in the analysis log.

# 21. Session information

After final analysis, record:

```{r session-info, eval=FALSE}
sessionInfo()
```

If `renv` is used, record the lockfile as well.

This helps distinguish a data change from a software-environment change.

# 22. Reporting candidate models

A candidate comparison table should identify:

* model family;
* number of observations;
* likelihood parameter count;
* RSS;
* RMSE;
* MAE;
* AIC;
* AICc;
* BIC;
* delta AICc;
* Akaike weight when calculated.

Do not report the best AICc model without also checking parameter plausibility and diagnostics [@ArchontoulisMiguez2015].

# 23. Reporting nonlinear traits

Biological traits often provide the most interpretable summary.

Depending on model family, useful quantities can include:

* asymptote;
* inflection time;
* response at inflection;
* maximum absolute growth rate;
* time of maximum absolute growth rate;
* time to 10 percent;
* time to 50 percent;
* time to 90 percent;
* area under the observed-support curve.

Report units.

Report uncertainty when it was estimated.

# 24. Reporting smooth trajectories

For smoothers, focus on:

* fitted trajectory;
* first derivative;
* second derivative when scientifically relevant;
* time support;
* uncertainty method if used;
* distinction between descriptive population smoothing and repeated-unit inference.

Regression smoothers have a long history in plant-growth analysis [@ShipleyHunt1996].

# 25. Reporting mixed models

For longitudinal mixed models, report:

* fixed-effects structure;
* random-effects structure;
* residual correlation structure;
* variance structure;
* estimation method;
* persistent unit;
* number of units;
* diagnostic findings.

Do not report repeated observations as if they were independent replicates [@PinheiroBates2000].

# 26. Reporting multiphase models

Report:

* number of phases;
* phase-ordering constraint;
* optimization strategy;
* complete-trajectory inflection or stability analysis;
* temporal support;
* whether alternative phase counts were considered.

For a sum of logistic phases, complete-trajectory critical points should not be replaced by component formulas [@MischanEtAl2015].

# 27. Reporting disturbance analyses

For defoliation or biomass-loss experiments, distinguish measured losses from model-estimated growth.

The iterative approach depends on observed losses of total biomass, leaf mass, and leaf area when those events occur [@AntenAckerly2001].

Do not reconstruct a loss event from a smooth curve and then present it as measured data.

# 28. Reporting Bayesian analyses

A minimal Bayesian report should include:

* model family;
* prior distributions;
* number of chains;
* iterations and warmup;
* backend;
* R-hat summary;
* effective sample sizes;
* divergences;
* posterior predictive checks;
* posterior trait summaries;
* PSIS-LOO diagnostics when model comparison is used.

Predictive stacking can be useful when several models remain plausible, but weights should be interpreted as predictive-combination weights rather than posterior model probabilities [@YaoEtAl2018].

# 29. Reporting bootstrap analyses

State:

* what was resampled;
* why that resampling unit matches the design;
* number of replicates;
* refitting success rate;
* confidence-interval type;
* seed;
* whether the interval summarizes coefficients or derived traits.

The resampling scheme is part of the method, not an implementation detail.

# 30. Reviewer bundle

For manuscript review, a useful bundle can contain:

1. analysis script;
2. data dictionary;
3. source-data checksum;
4. workflow RDS;
5. audit CSV;
6. trait table;
7. candidate comparison;
8. Markdown analysis report;
9. session information;
10. package tarball checksum.

This does not require that raw confidential data be publicly distributed.

Access restrictions can be documented separately.

# 31. Teaching bundle

For teaching, the bundle can be smaller:

* one simulated dataset;
* one analysis script;
* one planning-only workflow;
* one executed workflow;
* audit table;
* short interpretation exercise.

Students can compare what changes when the experimental unit is declared incorrectly.

# 32. Do not overwrite outputs silently

`growth_export()` defaults to `overwrite = FALSE`.

This is deliberate.

A reproducible workflow should not silently replace a previous result.

Use a new directory, versioned filename, or explicit overwrite decision.

# 33. File naming

Useful names include:

```text
sunflower_growth_workflow_2026-08-27.rds
sunflower_growth_audit_2026-08-27.csv
sunflower_growth_report_2026-08-27.md
```

For a formal release, prefer semantic version labels over dates alone.

# 34. Audit before export

A simple final sequence is:

```{r final-sequence, eval=FALSE}
w <- growth_workflow(...)

audit <- growth_workflow_audit(w)
print(audit)

stopifnot(!any(audit$status == "FAIL"))

growth_export(
  w,
  path = "outputs/agriGrowthFlow_bundle",
  format = "bundle"
)
```

This check is intentionally modest.

Model-specific diagnostics still have to be inspected before release.

# 35. Suggested Methods record

A reproducible Methods note can state:

*Analysis was performed with agriGrowthFlow 1.0.0. The persistent experimental unit and sampling mechanism were declared before fitting. The analysis objective and selected core strategy were stored in a consolidated workflow object. Model-specific diagnostics were reviewed before biological traits were interpreted. The final workflow was audited with `growth_workflow_audit()` and archived as an RDS object together with human-readable tables and a Markdown analysis report.*

Add the actual model family, uncertainty method, and experimental details.

# 36. Suggested data-availability record

If data can be shared:

*The analysis bundle contains the processed data, source checksum, complete agriGrowthFlow workflow object, analysis report, and session information required to reproduce the reported growth summaries.*

If data cannot be shared, state the restriction accurately.

Do not claim open availability when access is restricted.

# 37. Final audit checklist

Before sending results to a collaborator or reviewer:

- [ ] confirm the package version;
- [ ] confirm the source data file and checksum;
- [ ] confirm the persistent experimental unit;
- [ ] confirm the selected strategy;
- [ ] inspect model-specific diagnostics;
- [ ] inspect uncertainty diagnostics;
- [ ] verify units in traits and tables;
- [ ] inspect report wording for unsupported claims;
- [ ] save the complete RDS object;
- [ ] save human-readable tables;
- [ ] save session information;
- [ ] record the final package or tarball checksum.

# 38. Take-home message

Reproducibility requires more than saving the final table.

The 1.0.0 release separates three layers:

1. the complete computational object;
2. human-readable review outputs;
3. a transparent analysis contract.

Together, these layers make it easier to determine not only what result was obtained, but how that result was produced and what assumptions were declared.
